```{r echo=FALSE}
library(knitr)
```

---
title: "Damage Done by Severe Weather Events on Public Health and Economy in the United States"
author: "Thong B. Tran"
date: "Saturday, September 20, 2014"
output: html_document
---

##Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

In this project, we analyze the data recorded in the US to find out which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health and which types of events have the greatest economic consequences.

##Data Processing 
We load all the neccessary R libraries into RStudio.
```{r echo=TRUE}
library(plyr)
library(ggplot2)
require(gridExtra)
```

We download the data here "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"" into our working directory and load it into RStudio, then view its summary to understand the general information.
```{r echo=TRUE}
data <- read.csv("repdata-data-StormData.csv.bz2")
summary(data)
```

From the information, we get the number of observations in the data recorded from 1950 to 2011. Now, we create a new column named year and make a histogram of the collected data, then take a further step in processing the data by applying the ddplyr function to get the sums of fatalities, injuries, crop and property damages for each event type. Finally, we extract
ten highest values from those data and plot them. First, We process the fatality and injury data

```{r echo=TRUE}
health<- ddply(data, .(EVTYPE), summarize, INJURIES = sum(INJURIES),FATALITIES=sum(FATALITIES),na.rm=T)
data$year <- as.numeric(format(as.Date(data$BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), "%Y"))
hist(data$year,xlab="year",main="Data collected from 1950 to 2011",breaks=60)
fatalSub <- health[,c(1,3)]
injSub <- health[,1:2]
fatalities <- fatalSub[order(fatalSub$FATALITIES,decreasing = T),]
topTenFacilities <- head(fatalities, n=10)
injuries <- injSub[order(injSub$INJURIES,decreasing = T),]
topTenInjuries <- head(injuries, n= 10)
```

From the graph above, we see the lack of data from 1950 to 1994, and the highest amount of data available in 2011. Now, We need to process the PROPDMGEXP and CROPDMGEXP categories to calculate the damage done to the US in dollar. We do so by substitute B with 9, M with 6, K with 3, H with 2, and empty space with 0.

```{r echo=TRUE}
data$PROPDMGEXP <- as.character(data$PROPDMGEXP)
data$PROPDMGEXP[toupper(data$PROPDMGEXP) == "B"] <- "9"
data$PROPDMGEXP[toupper(data$PROPDMGEXP) == "M"] <- "6"
data$PROPDMGEXP[toupper(data$PROPDMGEXP) == "K"] <- "3"
data$PROPDMGEXP[toupper(data$PROPDMGEXP) == "H"] <- "2"
data$PROPDMGEXP <- as.numeric(data$PROPDMGEXP)
data$PROPDMGEXP[is.na(data$PROPDMGEXP)] <- 0
data$propertyDamage <- data$PROPDMG * 10^data$PROPDMGEXP

data$CROPDMGEXP <- as.character(data$CROPDMGEXP)
data$CROPDMGEXP[toupper(data$CROPDMGEXP) == "B"] <- "9"
data$CROPDMGEXP[toupper(data$CROPDMGEXP) == "M"] <- "6"
data$CROPDMGEXP[toupper(data$CROPDMGEXP) == "K"] <- "3"
data$CROPDMGEXP[toupper(data$CROPDMGEXP) == "H"] <- "2"
data$CROPDMGEXP[toupper(data$CROPDMGEXP) == ""] <- "0"
data$CROPDMGEXP <- as.numeric(data$CROPDMGEXP)
data$CROPDMGEXP[is.na(data$CROPDMGEXP)] <- 0
data$cropDamage <- data$CROPDMG * 10^data$CROPDMGEXP
```

Now, we can extract the ten highest values from the data after calculating the sum for each event type.
```{r echo=TRUE}
economy <- ddply(data, .(EVTYPE), summarize, propertyDamage = sum(propertyDamage),cropDamage=sum(cropDamage),na.rm=T)
property <- economy[order(economy$propertyDamage,decreasing = T),c(1,2)]
topTenPropertyDMG <- head(property, n=10)
crop <- economy[order(economy$cropDamage,decreasing = T),c(1,3)]
topTenCropDMG <- head(crop, n= 10)
```

##Result
Here, the result shows top ten harmful weather events with respect to population health
```{r echo=TRUE}
topTenInjuries
topTenFacilities
```

Based on the data above, we know that the most injury and fatality damages were caused by tornado. We make two plots to visualize these results
```{r echo=TRUE}
fatalityPlot <- qplot(EVTYPE, data = topTenFacilities, weight = FATALITIES, geom = "bar", binwidth = 2) + 
        scale_y_continuous("Fatalities") + 
        theme(axis.text.x = element_text(angle = 45, 
                                         hjust = 1)) + xlab("Severe Weather Type") + 
        ggtitle("Total Fatalities by Severe Weather\n Events in the U.S.\n from 1995 - 2011")
injuryPlot <- qplot(EVTYPE, data = topTenInjuries, weight = INJURIES, geom = "bar", binwidth = 2) + 
        scale_y_continuous("Injuries") + 
        theme(axis.text.x = element_text(angle = 45, 
                                         hjust = 1)) + xlab("Severe Weather Type") + 
        ggtitle("Total Injuries by Severe Weather\n Events in the U.S.\n from 1995 - 2011")
grid.arrange(fatalityPlot, injuryPlot, ncol = 2)
```
Here are the top ten events that have the greatest economic consequences
```{r echo=TRUE}
topTenPropertyDMG
topTenCropDMG
```
Based on the data above, we know that drought did the greatest damage to crop, and flood did the greatest damage to properties. We make two plots to visualize these results
```{r echo=TRUE}
cropPlot <- qplot(EVTYPE, data = topTenCropDMG, weight = cropDamage, geom = "bar", binwidth = 2) + 
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) + scale_y_continuous("Crop Damage") + 
        xlab("Severe Weather Type") + ggtitle("Total Crop Damage by \nSevere Weather Events in\n the U.S. from 1995 - 2011")

propertyPlot <- qplot(EVTYPE, data = topTenPropertyDMG, weight = propertyDamage, geom = "bar", binwidth = 2) + 
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) + scale_y_continuous("Property Damage")+ 
        xlab("Severe Weather Type") + ggtitle("Total Property Damage by\n Severe Weather Events in\n the U.S. from 1995 - 2011")
grid.arrange(cropPlot,propertyPlot, ncol=2)
```

